What We Eval in the Shadows

### 2021-04-16 paper/main.tex
Updates

From the \packageNbrunsRnd programs we capture \SEAll side effects from
\SEAllCalls \eval calls in \SEAllSites sites. Again, the challenge is to remove
the accidental side-effects caused by the R virtual machine implementation and
are not related to the user code. For example, the \c{.Random.seed} variable,
which contains the state of the random number generator is saved and restored
from and to the global environment every time the user calls one of the base
routines that use random numbers. Removing them leaves us with \SEUser side
effects (\SEUserRatio) from \SEUserCalls \eval calls (\SEUserCallsRatio) in
\SEUserSites sites (\SEUserSitesRatio). These sites are part of
\SEUserFunctions in \SEUserPackages packages. \SEFunsNighty functions are
responsible for 90\% of side effects. Half of the side-effects come just from
three functions: \c{plyr::allocate\_column} (allocates space for a new data
frame column), \c{withr::execute\_handlers} (executes deferred expressions), and
\c{foreach::doSEQ} (executes an expression on each element in a collection,
possibly in parallel).

Most of the \eval sites (\SESitesInEnvirRatio) do side effects in the
environment specified by the \c{envir} parameter (\cf
Section~\ref{sec:eval-in-r}), \SESitesNotInEnvirRatio modifies other
environments, and finally \SESitesBothEnvirRatio does both.
Table~\ref{tab:se-env} shows the class of the environment where \eval side
effects happen (\emph{target environment}). It follows the classification
described in Section~\ref{sec:env} with two extra classes: \emph{Object}, which
represents a side-effect that happens in S4 or R6 object environment, and
\emph{Local} that corresponds to a side effect in local variables of the \eval
caller. The table shows data for both \eval sites and functions. For a function
to have a given target environment class, all of its \eval sites that do side
effects must have the same target environment class The majority of the \eval
sites (\SESitesInOneClass) do all side effects consistently in one environment
class. The same happens on the function level. Almost half of the sites (over a
third of the functions) do side effects in either \emph{Local} or \emph{Object}
environments. This gives a ray of hope for the compiler. Even though it is
possible to do anything anywhere, the data suggest that most side-effects are
sane.

@@

% TODO type of variables
% TODO do assignments change types?

@@

In this paper we provide large-scale study of the use of the \eval function in
the R programming language. It is based primarily on a corpus of
\CranRunnableScripts scripts extracted from \CranPackages R packages. As
expected, \eval is used mainly in libraries and therefore we have found just a
few \eval call sites in scripts coming from Kaggle that we included to
represent code written by R practitioners. We used an instrumented version of
the R virtual machine, \rdyntrace, to capture all \eval calls traces
including the relevant side effects that happen within these calls. There is a
number of insights that can be learnt from the data.
%
\eval is widely used in the R core packages that come bundled with the
language. Because of that there is hardly any R code that would not use it.
\eval is used to implement some basic functionality such as package loading.
Putting the core packages aside, in CRAN, \eval is used by \PkgPackagesRatio
of the available \CranPackages packages.
%
The code executed by R \eval has unusual expressive power. It can essentially
modify anything anywhere which is wary for both R developers as well as R
compiler.
% TODO expressions passed to eval
% TODO side effects

% how much can we say about the threat model?

>>>>

From the \packageNbrunsRnd programs we capture \SEAll side effects from
\SEAllCalls \eval calls in \SEAllSites sites. Again, the challenge is to remove
the accidental side-effects caused by the R virtual machine implementation and
are not related to the user code. For example, the \c{.Random.seed} variable,
which contains the state of the random number generator is saved and restored
from and to the global environment every time the user calls one of the base
routines that use random numbers.
%
% \begin{wrapfigure}{r}{5cm} \hspace*{-20mm} \centering
% \includegraphics[width=90mm]{se-types.pdf} \caption{Distribution of R types
% in target environments for side effects.} \label{fig:se-types}
% \end{wrapfigure}
%
Removing them leaves us with \SEUser side effects (\SEUserRatio) from
\SEUserCalls \eval calls (\SEUserCallsRatio) in \SEUserSites sites
(\SEUserSitesRatio). These sites are part of \SEUserFunctions in
\SEUserPackages packages. \SEFunsNighty functions are responsible for 90\% of
side effects. Half of the side-effects come just from three functions:
\c{plyr::allocate\_column} (allocates space for a new data frame column),
\c{withr::execute\_handlers} (executes deferred expressions), and
\c{foreach::doSEQ} (executes an expression on each element in a collection,
possibly in parallel). Most of the \eval sites (\SESitesInEnvirRatio) do side
effects in the environment specified by the \c{envir} parameter (\cf
Section~\ref{sec:eval-in-r}), \SESitesNotInEnvirRatio modifies other
environments, and finally \SESitesBothEnvirRatio does both.
Table~\ref{tab:se-env} shows the class of the environment where \eval side
effects happen (\emph{target environment}). It follows the classification
described in Section~\ref{sec:env} with two extra classes: \emph{Object}, which
represents a side-effect that happens in S4 or R6 object environment, and
\emph{Local} that corresponds to a side effect in local variables of the \eval
caller. The table shows data for both \eval sites and functions. For a function
to have a given target environment class, all of its \eval sites that do side
effects must have the same target environment class The majority of the \eval
sites (\SESitesInOneClass) do all side effects consistently in one environment
class. The same happens on the function level. Almost half of the sites (over a
third of the functions) do side effects in either \emph{Local} or \emph{Object}
environments. This gives a ray of hope for the compiler. Even though it is
possible to do anything anywhere, the data suggest that most side-effects are
sane.

@@

Looking closely at the value types in variable updates, we observe that the
majority of \eval sites involve basic R vectors (\SEBasicTypeRatio) and lists
(\SEListTypeRatio). From the compiler perspective, we would like to know how
many sites changes function bindings. In the corpus, this happens in only
\SEClosureType sites, \SEClosureTypeLocal of them do that in the local
environment. We have manually inspected a few of these sites, but except for
manual injection of parameters into a \c{model.frame} execution environment, we
did not find a common use case.

% TODO do assignments change types?

@@

In this paper, we provide a large-scale study of the use of the \eval function
in the R programming language. It is based primarily on a corpus of
\CranRunnableScripts scripts extracted from \CranPackages R packages. As
expected, \eval is used mainly in libraries and therefore we have found just
a few \eval call sites in scripts coming from Kaggle that we included to
represent code written by R practitioners. We used an instrumented version of
the R virtual machine, \rdyntrace, to capture all \eval calls traces including
the relevant side effects that happen within these calls. Several insights can
be learnt from the data.
%
First, \eval is widely used in the R core packages that come bundled with the
language. Because of that, there is hardly any R code that would not use it.
\eval is used to implement some basic functionality such as package loading.
Putting the core packages aside, in CRAN, \eval is used by \PkgPackagesRatio of
the available \CranPackages packages.
%
The code executed by R \eval has unusual expressive power. It can essentially
modify anything anywhere which is wary for both R developers and R compiler.
% TODO expressions passed to eval
Even more so in the case of side effects that can quite literally happen
anywhere. The data however suggest that this is not the case. The majority of
eval sites do not do any observable side effects and when they do, it happens
in a predictable environment.

% how much can we say about the threat model?
Similarly to \citep{ecoop11}, we started looking into \evals in R with the
ambition to replace them by other feature that are more amenable to static
analysis. The data unfortunately suggests that replacing \eval is going to be
lengthly journey. Even if we disregard core packages as they are part of the
language and the \eval use is stable, in CRAN \eval is used widely and in more
sophisticated way than was the case in JavaScript. There the vast majority of
sites fitted an extremely simple categorization based on simple regular
expressions. Not so in R. While some usage, especially the one which calls
\eval with an expression parsed from string, can be categorized and often
replaced by equivalent yet more disciplined and safer code, the case of \eval
coming from reflection is much harder.
